---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Effective Pest Treament That Protects Pollinators"
subtitle: "https://github.com/shivanikuckreja/CitrolaKuckrejaSaltman_ENV872_EDA_FinalProject/tree/main/Project"
author: "Sam Saltman, Shivani Kuckreja, Jessica Citrola"
fontsize: 12pt
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()
# Load your packages
library(viridis)
library(tidyverse)
library(lubridate)
library(knitr)
library(ggplot2)
library(RColorBrewer)
library(pscl)


# Set your ggplot theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

# Load your datasets 

Original_ECOTOX_CSV <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
```


# Rationale and Research Questions

**Pollination is a critical component of agriculture. Bees are important pollinators. Our research looks to see if there are exposure methods and chemicals that do not cause significant harm to bees while eliminating pests. The goal of our research is to determine potential treatment methods that reduce pests while having little to no impact on bees.**

Questions:

1.   _Is there an exposure type that is more likely to cause mortality for bees vs. non-bee insects?_

2.   _Are there chemicals that are more likely to cause mortality for bees vs. non-bee insects?_
  

\newpage

# Dataset Information

Data Source: The dataset was pulled from a repository created for Environmental 
Data Analytics at Duke University in 2020. The data collected is from several 
EPA studies on neonicotinoids and their effects on insects. The data we will be 
analyzing is the type of chemical administered, how it was administered, and 
how both of these variables impact insects. 

In the wrangling process, we selected the relevant information to our topic. 
This includes the chemical type, chemical number, insect species, lifestage and age of the species, 
exposure type, the effect of the exposure and the measurement of the exposure. An example of an exposure type is giving food to a bee. An example of an effect is mortality, and the measurement is a more detailed analysis of the effect. 

We converted all these selected categorical variables to factors to prep for analysis. In the next step, we processed two data frames – oall bee species and all non-bee species. The split resulted in 2529 non-bee observations and 1407 bee observations. Lastly we added a mortality column using an ifelse statement. We did this to run a binomial glm in our analysis. We coded mortality as 1 and everything else as 0. 


```{r echo=FALSE}
Processed1_Filter <- Original_ECOTOX_CSV %>%
  select(CAS.Number, Chemical.Name, Species.Common.Name, Organism.Lifestage, Organism.Age, Exposure.Type, Effect, Effect.Measurement) 

#Making all columns categorical
Processed1_Filter$CAS.Number <- as.factor(Processed1_Filter$CAS.Number)
Processed1_Filter$Chemical.Name <- as.factor(Processed1_Filter$Chemical.Name)
Processed1_Filter$Species.Common.Name <- as.factor(Processed1_Filter$Species.Common.Name)
Processed1_Filter$Organism.Lifestage <- as.factor(Processed1_Filter$Organism.Lifestage)
Processed1_Filter$Organism.Age <- as.factor(Processed1_Filter$Organism.Age)
Processed1_Filter$Exposure.Type <- as.factor(Processed1_Filter$Exposure.Type)
Processed1_Filter$Effect <- as.factor(Processed1_Filter$Effect)
Processed1_Filter$Effect.Measurement <- as.factor(Processed1_Filter$Effect.Measurement)

#Processed1_Filter$CAS.Number <- as.factor(Processed1_Filter$CAS.Number)

write.csv(Processed1_Filter, row.names = FALSE, file = "./Data/Processed/ProcessedForColumns")

#Mortality <- Processed1_Filter %>%
  #filter(Effect == "Mortality")

 BeeMortalityExposureType <- Processed1_Filter %>%
  filter(Species.Common.Name == "Honey Bee" | Species.Common.Name == "Buff Tailed Bumblebee" | Species.Common.Name == "Carniolan Honey Bee" | Species.Common.Name == "Bumble Bee" | Species.Common.Name == "Italian Honeybee" | Species.Common.Name == "European Dark Bee" | Species.Common.Name == "Buff-tailed Bumblebee" | Species.Common.Name =="Stingless Bee"| Species.Common.Name == "Mason Bee")
 
 write.csv(BeeMortalityExposureType, row.names = FALSE, file = "./Data/Processed/BeeMortality")
 
  NonBeeMortalityExposure <- Processed1_Filter %>%
  filter(Species.Common.Name == "Japanese Beetle" | Species.Common.Name == "Wireworm" | Species.Common.Name == "Asian Citrus Psyllid" | Species.Common.Name == "Parasitoid Wasp" | Species.Common.Name =="Snout Beetle Family, Weevil" | Species.Common.Name =="Sweetpotato Whitefly" | Species.Common.Name =="Predatory Mite" | Species.Common.Name =="Scarab Beetle" | Species.Common.Name =="Ground Beetle Family" | Species.Common.Name =="Chalcid Wasp" | Species.Common.Name =="Spider/Mite Class" | Species.Common.Name =="Ladybird Beetle" | Species.Common.Name =="Argentine Ant" | Species.Common.Name =="Horned Oak Gall Wasp" | Species.Common.Name =="Tooth-necked Fungus Beetle" | Species.Common.Name =="Calico Scale" | Species.Common.Name =="Minute Parasitic Wasps" | Species.Common.Name =="Silkworm" | Species.Common.Name =="Moth And Butterfly Order" | Species.Common.Name =="Hemlock Wooly Adelgid" | Species.Common.Name =="Western Flower Thrips" | Species.Common.Name =="House Fly" | Species.Common.Name =="Spined Soldier Bug" | Species.Common.Name =="Eulophid Wasp" | Species.Common.Name =="Yellow Fever Mosquito" | Species.Common.Name =="Eastern Subterranean Termite" | Species.Common.Name =="Pea Aphid" | Species.Common.Name =="Glasshouse Potato Wasp" | Species.Common.Name =="Two Spotted Lady Beetle" | Species.Common.Name =="(Other)" | Species.Common.Name =="Parasitic Wasp" | Species.Common.Name =="Asian Lady Beetle" | Species.Common.Name =="Parastic Wasp" | Species.Common.Name =="Erythrina Gall Wasp" | Species.Common.Name =="Sevenspotted Lady Beetle" | Species.Common.Name =="Aphid Family" | Species.Common.Name =="Braconid Wasp" | Species.Common.Name =="Ladybird Beetle Family" | Species.Common.Name =="Spring Tiphia" | Species.Common.Name =="Rove Beetle Family" | Species.Common.Name =="Convergent Lady Beetle" | Species.Common.Name =="Tobacco Flea Beetle" | Species.Common.Name =="Beetle" | Species.Common.Name =="Leaf Beetle Family" |Species.Common.Name =="Codling Moth" | Species.Common.Name =="Fairyfly Parasitoid" | Species.Common.Name =="Mirid Bug" | Species.Common.Name =="Vedalia Beetle" | Species.Common.Name =="Egg Parasitoid" | Species.Common.Name =="Oystershell Scale Parasitoid" | Species.Common.Name =="Mite" | Species.Common.Name =="Corn Earworm" |Species.Common.Name =="Ox Beetle" | Species.Common.Name =="Armoured Scale Family" | Species.Common.Name =="Monarch Butterfly" | Species.Common.Name =="Braconid Parasitoid" | Species.Common.Name =="Jassid" | Species.Common.Name =="Pond Wolf Spider" | Species.Common.Name =="Lacewing" | Species.Common.Name =="Ant Family" |Species.Common.Name =="Euonymus Scale" | Species.Common.Name =="Minute Pirate Bug" | Species.Common.Name =="Colorado Potato Beetle" | Species.Common.Name =="Beetle Order" | Species.Common.Name =="True Bug Order" | Species.Common.Name =="Cotton Aphid" | Species.Common.Name =="" | Species.Common.Name =="" |Species.Common.Name =="Thrip Order" | Species.Common.Name =="Cabbage Looper" | Species.Common.Name =="Parasitoid" | Species.Common.Name =="Tobacco Aphid" | Species.Common.Name =="Citrus Leafminer" | Species.Common.Name =="Mosquito" | Species.Common.Name =="Flatheaded Appletree Borer" | Species.Common.Name =="Potato Leafhopper" |Species.Common.Name =="Black-spotted Lady Beetle" | Species.Common.Name =="Lady Beetle" | Species.Common.Name =="Mulberry Pyralid" | Species.Common.Name =="Araneoid Spider Order" | Species.Common.Name =="Insect Class" | Species.Common.Name =="Hemlock Woolly Adelgid Lady Beetle" | Species.Common.Name =="Onion Thrip" | Species.Common.Name =="Green Peach Aphid" |Species.Common.Name =="Red Scale Parasite" | Species.Common.Name =="Diamondback Moth" | Species.Common.Name =="Predatory Bug" | Species.Common.Name =="Common Thrip" | Species.Common.Name =="Mite Order" | Species.Common.Name =="Spotless Ladybird Beetle" | Species.Common.Name =="Southern House Mosquito" | Species.Common.Name =="Apple Maggot")
  
 write.csv(NonBeeMortalityExposure, row.names = FALSE, file = "./Data/Processed/NonBeeMortality")

```



Detail | Description
------- | ---------
Data Source | EPA ECOTOX Knowledgebase
------- | ---------
Retrieved From | https://cfpub.epa.gov/ecotox/help.cfm
------- | ---------
Variables Used | Chemical Number, Chemical Name, Species Common Name, Organism Lifestage, Organism Age, Exposure Type, Effect, Effect Measurement
------- | ---------
Date Range | 1982-2019
\newpage

# Exploratory Analysis 

Summary of all species in study
```{r echo=FALSE, tab.cap = "Species List - Sample Number"}

kable(summary(Processed1_Filter$Species.Common.Name), caption = "Species List")
```

Bar chart comparing exposure type to mortality count of bees. It initially looked like bees are more likely to die when chemical exposure comes from consuming food.The other exposure types do not look particularly significant

```{r echo=FALSE,fig.cap = "Bee Mortality by Exposure Type", warning=FALSE}

GGPlotMortalityBeeFilter <- BeeMortalityExposureType %>%
  filter(Effect == "Mortality")
 
GGPlotMortalityNonBeeFilter <- NonBeeMortalityExposure %>%
  filter(Effect == "Mortality")

#reorder or order?? Tried but not working
 GGPlot_Bee_Mortality_Exposure <- ggplot(GGPlotMortalityBeeFilter) +
   aes(x = Exposure.Type) +
   geom_bar(aes(color = Exposure.Type, fill = Exposure.Type), method = "lm") +
   scale_fill_viridis(discrete = TRUE, option = "B") +
   xlab("Exposure Type") +
   ylab("Mortality Count") +
   theme(axis.text.x = element_text(angle = 90)) 
 print(GGPlot_Bee_Mortality_Exposure)
 
```

Bar chart comparing exposure type to mortality count of non-bees. Exposure from an environmental source appears more lethal to non-bees species. 

```{r echo=FALSE, fig.cap = "Non-bee Mortality by Exposure Type", warning=FALSE}
 
 
 GGPlot_NonBee_Mortality_Exposure <- ggplot(GGPlotMortalityNonBeeFilter) +
   aes(x = Exposure.Type) +
   geom_bar(aes(color = Exposure.Type, fill = Exposure.Type), method = "lm") +
   scale_fill_viridis(discrete = TRUE, option = "B") +
   ylab("Mortality Count") +
   xlab("Exposure Type") +
   theme(axis.text.x = element_text(angle = 90))
 print(GGPlot_NonBee_Mortality_Exposure)
```

The next area of exploration was looking into how many bee and non-bee samples died from exposure to certain chemical compounds. We used the chemical number as the chemical names are complex.

This bar chart looking at bees suggests that three types of chemical compounds in this dataset are toxic to bees.

```{r echo=FALSE, fig.cap = "Bee Mortality by Chemical", warning=FALSE}
 
GGPlot_Bee_Mortality_Chemical <- ggplot(GGPlotMortalityBeeFilter) +
   aes(x = CAS.Number) +
   geom_bar(aes(color = CAS.Number, fill = CAS.Number), method = "lm") +
   scale_fill_viridis(discrete = TRUE, option = "B") +
   xlab("Chemical") +
   ylab("Mortality Count") +
   theme(axis.text.x = element_text(angle = 90))
 print( GGPlot_Bee_Mortality_Chemical)
```

This bar chart looking at bees suggests that three types of chemical compounds in this dataset are toxic to non-bees.

```{r echo=FALSE, fig.cap = "Non-bee Mortality by Chemical", warning=FALSE}
 
GGPlot_NonBee_Mortality_Chemical <- ggplot(GGPlotMortalityNonBeeFilter) +
   aes(x = CAS.Number) +
   geom_bar(aes(color = CAS.Number, fill = CAS.Number), method = "lm") +
   scale_fill_viridis(discrete = TRUE, option = "B") +
   ylab("Mortality Count") +
   xlab("Chemical") +
   theme(axis.text.x = element_text(angle = 90))
 print(GGPlot_NonBee_Mortality_Chemical)
```

Next we analyze to see if there is any statistical significance supporting these observations.
\newpage

# Analysis



## Question 1: Is there an exposure type that is more likely to cause mortality for bees vs. non-bee insects?

```{r}

#Which exposure types have a significant effect (mortality) on bee species?
BeeMortalityExposureType$Mortality <- ifelse(BeeMortalityExposureType$Effect=="Mortality", 1, 0) 

#Making new column categorical
BeeMortalityExposureType$Mortality <- as.factor(BeeMortalityExposureType$Mortality)

logit <- glm(Mortality ~ Exposure.Type, data = BeeMortalityExposureType, family = "binomial")
summary(logit)

#For every one unit change in Topical, the odds of mortality increase by 1.1787. The P value of topical exposure type is 0.009, and is statistically significant.
#Significance: Total topical samples 99 - mortality - 65
#Additional data collection is needed on: Exposure through soil contact, drinking

#Pseudo R2
 pR2(logit)

```

```{r}

#Which exposure types have a significant effect (mortality) on non bee species?

NonBeeMortalityExposure$Mortality <- ifelse(NonBeeMortalityExposure$Effect=="Mortality", 1, 0) 

#Making new column categorical
NonBeeMortalityExposure$Mortality <- as.factor(NonBeeMortalityExposure$Mortality)

logit2 <- glm(Mortality ~ Exposure.Type, data = NonBeeMortalityExposure, family = "binomial")
summary(logit2)

#Dipped or soaked, Direct application, Environmental- unspecified, Foliar spray, Ground spray, Hand spray, Spray, and Topical, general are all significant 
#Plenty of samples for non-bees - correlations  
#Conclusion: Spray could be an effective anthropogenic exposure technique to eliminate pests and preserve bees. Direct application like topical techniques should be avoided as they significantly harm all species. There appears to be environmental exposure types that can promote desired effect as well.

#Pseudo R2
 pR2(logit2)


```


## Question 2: Are there chemicals that are more likely to cause mortality for bees vs. non-bee insects?
```{r}
#Which chemical types have a significant effect (mortality) on bee species?

logit3 <- glm(Mortality ~ CAS.Number, data = BeeMortalityExposureType, family = "binomial")
summary(logit3)

 #Need more data on 58842209 105843365 and 58842209, 150824478
 # Every other chemical is statistically significant - 111988499 135410207 138261413  153719234 165252700 210880925 

#Pseudo R2
 pR2(logit3)
```

```{r}
#Which exposure types have a significant effect (mortality) on non bee species?

logit4 <- glm(Mortality ~ CAS.Number, data = NonBeeMortalityExposure, family = "binomial")
summary(logit4)

#There is no chemical that is effective at eliminating all non-bee species. 
#Conclusion - exposure mechanism of chemical is a more reliable factor at understanding mortality than chemical itself. Bees do not respond well to any of the adminstered chemicals. Any treatment should focus on using the chemicals in ways that are ineffective at exposing bees and effective at exposing nonbees

#Pseudo R2
 pR2(logit4)
```


\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
